Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-11 paper/ResponseToReviews.tex

- Fig 2: Normalizing high versus low DI stocks based on equal TCID50 to compare IFN induction is not appropriate since these stocks will differ in total particle number. In this experiment, the cells would receive a much higher overall dose of viral RNAs with the high DI stock compared to the others.

{\color{red}
Gently explain why this is not relevant for the point we're trying to make.
}

- Fig. 3B: While I appreciate the density of information presented by this figure, the authors' main point concerning heterogeneity in the viral proportion of total mRNA reads might be made more clearly by simply plotting the relative proportion of mRNAs that are viral per cell.

{\color{black}
The reviewer points out that Figure 3B is a complicated way to show the heterogeneity in total mRNA.
In response to this comment, we have also added a more straightforward presentation in Figure 3 - Figure supplement 1, which shows the distribution of the viral proportion as a cumulative fraction plot.
This new plot should address the reviewer's point.

Incidentally, the most obvious way to plot these data would initially simply seem to be a histogram or kernel density plot with fraction of mRNA from virus on the x-axis, and proportion of cells on the y-axis.
However, such a plot turns out to be visually un-interpretable, because so much of the density is at zero (un-infected cells) that it is impossible to see any information for other parts of the plot for any reasonable y-axis scaling. 
This problem can be remedied by \emph{only} showing infected cells (getting rid of the huge bump in density at zero due to un-infected cells) -- but that can only be done after we have introduced a criterion to define which cells are infected, which doesn't happen until later in the text (and we want to show all the data before this censoring).
Therefore, we also plot a more straightforward histogram display in Figure 4E for only the infected cells after censoring un-infected cells.
We believe the combination of Figure 3B, the new Figure 3 - Figure - supplement 1, and Figure 4E effectively capture all the effects given these complexities of visual display for data distributed in such an asymmetric way.
}

### 2018-01-11 paper/paper.tex
added requested supplement to Fig 3 to show viral mRNA per cell distribution in another way

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing HA protein.
A549 cells were infected at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}

\subsection{Single cells show an extremely wide range of expression of viral mRNA.}
We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate a 8-hours.
We replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
The exception was the replicate 8-hour sample, for which we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}B).

\begin{figure}
\centerline{\includegraphics[width=\linewidth]{figures/p_cell_mRNA_summary.pdf}}
\caption{\label{fig:cells}
There is a very wide distribution in the amount of viral mRNA per cell.
{\bf (A)} 
Number of cells sequenced for each sample.
{\bf (B)} 
The number of cellular and viral mRNAs detected for each cell is plotted as a point.
The blue lines show the overall distribution of the number of cellular mRNAs per cell.
The orange rug plot at the left of each panel shows the distribution of the number of viral mRNAs per cell.
Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
}
\end{figure}

>>>>

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing HA protein.
A549 cells were infected at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}
\end{figure}

\subsection{Single cells show an extremely wide range of expression of viral mRNA.}
We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate a 8-hours.
We replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
The exception was the replicate 8-hour sample, for which we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}B).

\begin{figure}
\centerline{\includegraphics[width=\linewidth]{figures/p_cell_mRNA_summary.pdf}}
\caption{\label{fig:cells}
There is a very wide distribution in the amount of viral mRNA per cell.
{\bf (A)} 
Number of cells sequenced for each sample.
{\bf (B)} 
The number of cellular and viral mRNAs detected for each cell is plotted as a point.
The blue lines show the overall distribution of the number of cellular mRNAs per cell.
The orange rug plot at the left of each panel shows the distribution of the number of viral mRNAs per cell.
Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac} shows the exact distributions of the fraction of viral mRNA per cell.
}
\figsupp[Cumulative fraction plot of proportion of total mRNA from virus. \label{figsupp:viral_mRNA_cumulfrac}]{
For each sample, this plot shows the fraction of all cells that derive at least the indication fraction of their mRNA from influenza virus.
}{\includegraphics[width=0.5\linewidth]{figures/p_mRNA_per_cell_cumul.pdf}}
\end{figure}
